Elephant Seal Age JAE Intraspecific Variation

### 2024-05-29 paper/paperv3.qmd
Update pup surv/recr

```{r}
#| label: fig-raw-data

@@

summary(breed_age_mod)

@@

```{r}
#| label: fig-breeding-mod 
#| fig-width: 7.5

@@

##NOTE: Need to take out the instances where recruitment > survival
#what does ggpredict do with random effects? 
#thin gray lines for the different year effects? show the interannual variation (5 that include extreme ones, just something to illustrate the variation of year)

# Isolate pup survival/recruitment 
surv_recruit_data <- sealdat %>% 
  drop_na(pupID) %>% 
  # Only pups born 2016 or earlier (otherwise survival biiased)
  filter(pupyearborn <= 2016) %>% 
  mutate(pup_survived_int = ifelse(pup_survived, 1L, 0L)) %>% 
  # Only keep maternal age classes with 5+ births, which drops moms 19+
  group_by(age) %>% 
  mutate(n_pups = n()) %>% 
  ungroup() %>%
  filter(n_pups >= 5)

# Fit pup *survival* models pre- and post-senescence
# Commented-out formulae vary age10 coef by pupsex, end up with similar
# results. Can add to supplemental?
pup_surv_model1 <- glmer(
  # pup_survived_int ~ age10 : pupsex + (1 | animalID) + (1 | year),
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age < age_senesce),
  family = "binomial"
)
pup_surv_model2 <- glmer(
  # pup_survived_int ~ age10 : pupsex + (1 | animalID) + (1 | year),
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age >= age_senesce),
  family = "binomial"
)

# Fit pup *recruitment* models pre- and post-senescence
pup_recr_model1 <- glmer(
  puprecruited ~ age10 + (1 | animalID) + (1 | year),
  # For recruitment, ONLY female pups
  data = filter(surv_recruit_data, 
                age < age_senesce,
                pupsex == "F"),
  family = "binomial"
)
pup_recr_model2 <- glmer(
  puprecruited ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, 
                age >= age_senesce,
                pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

# Summary table
surv_recr_summ <- tibble(
  response = rep(c("Survival", "Recruitment"), each = 2),
  n = map_dbl(
    list(pup_surv_model1@frame, 
         pup_surv_model2@frame, 
         pup_recr_model1@frame, 
         pup_recr_model2@frame),
    nrow
  ),
  model = rep(c("Pre-senescence", "Post-senescence"), 2),
  age_coef = c(fixef(pup_surv_model1)["age10"],
               fixef(pup_surv_model2)["age10"],
               fixef(pup_recr_model1)["age10"],
               fixef(pup_recr_model2)["age10"]),
  age_coef_se = c(summary(pup_surv_model1)$coef["age10", "Std. Error"],
                  summary(pup_surv_model2)$coef["age10", "Std. Error"],
                  summary(pup_recr_model1)$coef["age10", "Std. Error"],
                  summary(pup_recr_model2)$coef["age10", "Std. Error"]),
  `p(coef<0)` = c(pnorm(summary(pup_surv_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_surv_model2)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model2)$coef["age10", "z value"]))
)

surv_recr_summ %>% 
  mutate(across(where(is.numeric), \(x) signif(x, 3))) %>% 
  knitr::kable()

@@

```{r}
#| label: fig-repro-age
#| fig-cap: "Offspring survival decreases as mothers senesce. **A** Survival of female and male offspring (filled points) is not negatively impacted by maternal age (blue line) prior to age 12 (vertical dashed line). After age 12, offspring survival declines with maternal age (red line), i.e., maternal effect senescence. Hollow points represent female offspring recruitment, which did not significantly decline with maternal age after age 12. **B** The coefficient of age w.r.t. offspring survival (solid points) was not significantly negative before age 12 (blue), but was significantly negative after (red). The coefficient of age w.r.t. female offspring recruitment (hollow points) was not significant in either maternal age class. Lines represent standard error of the coefficient, which is ommitted for the post-age 12 offspring recruitment coefficent (hollow red point) because the standard error was too large (14.8) to visually represent without hiding other patterns. Numbers at top indicate sample size in each model."
#| fig-width: 5.5

#B Pup survival and recruitment doesn't change with increased maternal age.

# Pup survival 
surv_pred <- rbind(
  # Pre-senescence
  ggpredict(
    pup_surv_model1,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "young"),
  # Post-senescence
  ggpredict(
    pup_surv_model2,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "old")
)

repro_raw <- surv_recruit_data %>% 
  # Uncomment to group by female and male offspring
  # mutate(
  #   pupsex = factor(pupsex, 
  #                   labels = c("Female offspring", "Male offspring"))
  # ) %>% 
  # group_by(age, pupsex) %>% 
  group_by(age) %>% 
  summarize(n_born = n(), 
            n_surv = sum(pup_survived_int),
            perc_surv = n_surv / n_born, 
            se_surv = (perc_surv * (1 - perc_surv) / n_born)^0.5,
            f_born = sum(pupsex == "F"),
            f_recr = sum(puprecruited & pupsex == "F"),
            perc_recr = f_recr / f_born,
            se_recr = (perc_recr * (1 - perc_recr) / f_born)^0.5,
            .groups = "drop")

surv_recr_fig <- ggplot(surv_pred, aes(x = age)) +
  # Survival by age (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Raw survival and recruitment data
  geom_pointrange(aes(y = perc_surv,
                      ymin = perc_surv - se_surv,
                      ymax = perc_surv + se_surv),
                  repro_raw) +
  geom_pointrange(aes(y = perc_recr,
                      ymin = perc_recr - se_recr,
                      ymax = perc_recr + se_recr),
                  repro_raw,
                  shape = 21) +
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_x_continuous("Age", limits = c(3.5, 18.5)) +
  scale_y_continuous("Offspring survival/recruitment", 
                     limits = c(0, 0.65), 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  # Uncomment to facet by offspring sex
  # facet_wrap(~pupsex) +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))

# Survival/recruitment figure. In the caption, mention the post-senescence
# recruitment std err was omitted because it was 14.8, so it swamps all the
# other std errs
surv_recr_coef_fig <- surv_recr_summ %>% 
  mutate(response = fct_rev(str_sub(response, 1, 4)),
         model = fct_rev(str_extract(model, ".*-")),
         # Remove post-senescence recruitment std err
         age_coef_se = c(age_coef_se[1:3], 0)) %>% 
  ggplot(aes(x = response, ymax = 1.1)) +
  geom_pointrange(aes(ymin = age_coef - age_coef_se, 
                      y = age_coef,
                      ymax = age_coef + age_coef_se,
                      color = model,
                      shape = response),
                  position = position_dodge(width = 0.9)) +
  geom_text(aes(label = n, group = model), 
            y = 0.9, 
            size = 3,
            angle = 30,
            hjust = 0,
            position = position_dodge(width = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_shape_manual(values = c(16, 1)) +
  labs(x = "Response", y = "Age coefficient") +
  expand_limits(y = 1.1) +
  theme(legend.position = "none")

(surv_recr_fig | surv_recr_coef_fig) + 
  plot_layout(widths = c(3, 1)) +
  plot_annotation(tag_levels = "A")

@@

Sample size n = `r nrow(surv_recruit_data)`.

@@

ggplot(breed_pred_seg, aes(age, predicted)) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat)) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_fill_manual(values = c("cornflowerblue", "firebrick")) +
  scale_y_continuous("Breeding", labels = scales::percent) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

>>>>

```{r}
#| label: fig-raw-data
#| fig-width: 9

@@

summary(breed_age_mod)$coefficients

@@

```{r}
#| label: fig-breeding-mod

@@

##NOTE: Need to take out the instances where recruitment > survival

# Isolate pup survival/recruitment 
surv_recr_data <- sealdat %>% 
  drop_na(pupID) %>% 
  # Only pups born 2016 or earlier (otherwise survival biased)
  filter(pupyearborn <= 2016) %>% 
  mutate(pup_surv_int = as.integer(pup_survived),
         pup_recr_int = as.integer(puprecruited)) %>% 
  # Only keep maternal age classes with 5+ births, which drops moms 19+
  group_by(age) %>% 
  mutate(n_pups = n()) %>% 
  ungroup() %>%
  filter(n_pups >= 5)

# Fit pup *survival* model
pup_surv_mod <- glmer(
  pup_surv_int ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year_fct),
  surv_recr_data,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

# Fit pup *recruitment* model (female offspring only)
pup_recr_mod <- glmer(
  pup_recr_int ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year_fct),
  filter(surv_recr_data, pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

"Survival"
summary(pup_surv_mod)$coefficients
"Recruitment"
summary(pup_recr_mod)$coefficients

@@

```{r}
#| label: fig-repro-age
#| fig-width: 9

# Summarize the raw data
surv_recr_summ <- surv_recr_data %>% 
  group_by(age) %>% 
  summarize(n_pups = n(), 
            n_fpups = sum(pupsex == "F"),
            perc_surv = sum(pup_surv_int) / n_pups, 
            se_surv = (perc_surv * (1 - perc_surv) / n_pups)^0.5,
            perc_recr = sum(pup_recr_int) / n_fpups, 
            se_recr = (perc_recr * (1 - perc_recr) / n_fpups)^0.5) %>% 
  filter(age <= 20)

# Create pup survival POPULATION-LEVEL (i.e. without rand effects) predictions
surv_pred_pop <- ggpredict(
  pup_surv_mod, 
  terms = "age10 [all]",
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10) %>% 
  filter(age <= 20)

# Create pup survival WEIGHTED-AVERAGE predictions using random effects
seal_years <- count(sealdat, year_fct)
surv_pred_wgt <- ggpredict(
  pup_surv_mod,
  terms = c("age10 [all]", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         year_fct = group) %>% 
  filter(age <= 20) %>% 
  left_join(seal_years, by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(predicted = weighted.mean(predicted, n))

fig_surv <- ggplot(surv_pred_pop, aes(age, predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(linewidth = 1.2, color = "grey50") +
  geom_line(data = surv_pred_wgt, 
            color = "cornflowerblue", 
            linewidth = 1.2) +
  geom_pointrange(aes(y = perc_surv, 
                      ymin = perc_surv - se_surv, 
                      ymax = perc_surv + se_surv),
                  surv_recr_summ) +
  scale_y_continuous("Offspring survival", 
                     labels = scales::percent,
                     limits = c(0, 0.62)) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

# Create pup recruit POPULATION-LEVEL (i.e. without rand effects) predictions
recr_pred_pop <- ggpredict(
  pup_recr_mod, 
  terms = "age10 [all]",
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10) %>% 
  filter(age <= 20)

# Create pup recruit WEIGHTED-AVERAGE predictions using random effects
seal_years <- count(sealdat, year_fct)
recr_pred_wgt <- ggpredict(
  pup_recr_mod,
  terms = c("age10 [all]", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         year_fct = group) %>% 
  filter(age <= 20) %>% 
  left_join(seal_years, by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(predicted = weighted.mean(predicted, n))

fig_recr <- ggplot(recr_pred_pop, aes(age, predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(linewidth = 1.2, color = "grey50") +
  geom_line(data = recr_pred_wgt, 
            color = "cornflowerblue", 
            linewidth = 1.2) +
  geom_pointrange(aes(y = perc_recr, 
                      ymin = perc_recr - se_recr, 
                      ymax = perc_recr + se_recr),
                  surv_recr_summ) +
  scale_y_continuous("Offspring recruitment", 
                     labels = scales::percent,
                     limits = c(0, 0.62)) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

(fig_surv | fig_recr) + plot_annotation(tag_levels = "A")

@@

Sample size n = `r nrow(surv_recr_data)`.

@@

ggplot(breed_pred_seg, aes(age, predicted)) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat), linewidth = 1.2) +
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_fill_manual(values = c("cornflowerblue", "firebrick")) +
  scale_y_continuous("Breeding", labels = scales::percent) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

```{r}
#| label: seg-repro-mod

# Fit pup *survival* model
pup_surv_seg <- glmer(
  pup_surv_int ~ age10 * age_cat + (1 | animalID) + (1 | year),
  surv_recr_data,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

# Fit pup *recruitment* model
pup_recr_seg <- glmer(
  pup_recr_int ~ age10 * age_cat + (1 | animalID) + (1 | year),
  filter(surv_recr_data, pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

"Survival"
summary(pup_surv_seg)$coefficients
"Recruitment"
summary(pup_recr_seg)$coefficients

```{r}
#| label: fig-seg-repro-mod
#| fig-width: 9

surv_pred_seg <- ggpredict(pup_surv_seg, 
                           terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce))

fig_surv <- ggplot(surv_pred_seg, aes(age, predicted)) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat), linewidth = 1.2) +
  geom_pointrange(aes(y = perc_surv, 
                      ymin = perc_surv - se_surv, 
                      ymax = perc_surv + se_surv),
                  surv_recr_summ) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_fill_manual(values = c("cornflowerblue", "firebrick")) +
  scale_y_continuous("Offspring survival", labels = scales::percent) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

recr_pred_seg <- ggpredict(pup_recr_seg, 
                           terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce))

fig_recr <- ggplot(recr_pred_seg, aes(age, predicted)) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat), linewidth = 1.2) +
  geom_pointrange(aes(y = perc_recr, 
                      ymin = perc_recr - se_recr, 
                      ymax = perc_recr + se_recr),
                  surv_recr_summ) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_fill_manual(values = c("cornflowerblue", "firebrick")) +
  scale_y_continuous("Offspring recruitment", labels = scales::percent) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

(fig_surv | fig_recr) + plot_annotation(tag_levels = "A")
